---
title: Data Structures
teaching: 60
exercises: 25
source: Rmd
---

::::::::::::::::::::::::::::::::::::::: objectives

- To begin exploring data frames, and be able to visualize them in Rstudio.
- To be able to ask questions from R about the type, class, and structure of an object.
- To understand how to add rows or columns to a data frame.
- To understand how to subset parts of data frames.

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- How can I read data in R?
- What are the basic data types in R?
- How can I manipulate a data frame?

::::::::::::::::::::::::::::::::::::::::::::::::::

```{r, include=FALSE}
options(stringsAsFactors = FALSE)
cats_orig <- data.frame(coat = c("calico", "black", "tabby"), weight = c(2.1, 5, 3.2), likes_catnip = c(1, 0, 1), stringsAsFactors = FALSE)
cats_bad <- data.frame(coat = c("calico", "black", "tabby", "tabby"), weight = c(2.1, 5, 3.2, "2.3 or 2.4"), likes_catnip = c(1, 0, 1, 1), stringsAsFactors = FALSE)
cats <- cats_orig
```

One of R's most powerful features is its ability to deal with tabular data -
such as you may already have in a spreadsheet or a CSV file. Let's start by
making a toy dataset in your `data/` directory, called `feline-data.csv`:

```{r}
cats <- data.frame(coat = c("calico", "black", "tabby"),
                    weight = c(2.1, 5.0, 3.2),
                    likes_catnip = c(1, 0, 1))
```

Let's now view the cats dataframe. We can to this a number of different ways

```{r}
cats

View(cats)
```

We can now ask R about the structure of our dataframe — variables (columns) may be numbers, characters, or factors

```{r}
str(cats)
```

Imagine we were interested in asking how whether or not cats like catnip affect other metrics like their weight.
In this case, we would need to set "likes_catnip" to a factor level. These are categorical variables that only take on 
finite, discrete values.

Let's start by isolating the "likes_catnip" column visually using the $ function.

```{r}
cats$likes_catnip
```

Now, we can create a fourth column in the cats dataframe that contains this information as a factor, and view the structure of the
dataset again.

```{r}
cats$factor_catnip <-as.factor(cats$likes_catnip)

cats
View(cats)
str(cats)
```

Finally, let's add a fourth piece of information to our data frame as a fifth column.

```{r}
age <- c(2, 3, 5)
cats
```

We can then add this as a column via:

```{r}
cbind(cats, age)
```

Note that if we tried to add a vector of ages with a different number of entries than the number of rows in the data frame, it would fail:

```{r, error=TRUE}
age <- c(2, 3, 5, 12)
cbind(cats, age)

age <- c(2, 3)
cbind(cats, age)
```

Why didn't this work? Of course, R wants to see one element in our new column
for every row in the table:

```{r}
nrow(cats)
length(age)
```

So for it to work we need to have `nrow(cats)` = `length(age)`. Let's overwrite the content of cats with our new data frame.

```{r}
age <- c(2, 3, 5)
cats <- cbind(cats, age)

cats
```

Now how about adding rows? We already know that the rows of a
data frame are lists:

```{r}
newRow <- list("tortoiseshell", 3.3, TRUE, 1, 9)
cats <- rbind(cats, newRow)
```

Let's confirm that our new row was added correctly. 

```{r}
cats
```

We can now save `cats` as a CSV file. It is good practice to call the argument
names explicitly so the function knows what default values you are changing. Here we
are setting `row.names = FALSE`. Recall you can use `?write.csv` to pull
up the help file to check out the argument names and their default values.

##################################################################################################

NOTE TO SELF : this is not presently working. I can get this to work for the github just fine, but will need to coordinate with Amanda on
the file directory we are expecting students to set up

##################################################################################################

```{r}
write.csv(cats, "data/feline-data.csv")
```

Okay, so 
